Density-dependent selection and the limits of relative fitness

### 2018-01-02 manuscript.tex
l

The second justification, which primarily grew out of a controversy over Haldane's ``cost of selection'', is to appeal to the existence of a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population,kimura1969natural,nei1971fertility}. Selection can then be concentrated at the juvenile phase, uncoupling selection from population density at the adult phase unless it is so strong that the reproductive excess is depleted. This justifies Eq.~\eqref{eq:canonical} because, for a population in demographic equilibrium, selective sweeps do not affect density, and so the density-dependence of selection does not matter. Unfortuantely this reproductive excess literature is also poorly integrated with population ecology. \cite{kimura1969natural} took constant $N$ as a requirement and then derived some variants of the logistic model that satisfy this requirement. \cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types. As a result, the role of reproductive excesses in justifying Eq.~\eqref{eq:canonical} is still largely verbal.

@@

To briefly review: based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme proposed a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) \citep{macarthur_1962}, with the latter taken to mean selection for greater saturation density \citep{gill_1974}. A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}. Setting aside $r$ which confers no selective advantage at equilibrium, we are left with $K$ and $\alpha$, which both behave like $\delta$ in that they are density-dependent and cause density to change over a sweep (although $N$ only dips temporarily during an $\alpha$-sweep). Thus, strong selection is sufficient for relative fitness to break down in the classical view of density-dependent selection. Indeed, in the defense of Eq.~\eqref{eq:canonical} given by \cite{kimura1969natural}, it was assumed that $s$ will be a few percent at most. While this may be reasonable for adaptive mutations, there is no reason to expect selection on standing variation to be so weak, wild \textit{Drosophila} being an obvious counter-example \citep{bergland_2014}.

Our variable-density lottery model shows that it is not simply a lack of ecological realism that underlies the contrast between relative fitness with the classical view of density-dependent selection. Rather, in many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g. the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation). As noted in the introduction, this precludes selection concentrated at a fragile juvenile stage as a result of a reproductive excess \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}.

>>>>

The second justification, which primarily grew out of a controversy over Haldane's ``cost of selection'', is to appeal to the existence of a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population,kimura1969natural,nei1971fertility}. Selection can then be concentrated at the juvenile phase, uncoupling selection from population density at the adult phase unless it is so strong that the reproductive excess is depleted. This justifies Eq.~\eqref{eq:canonical} because, for a population in demographic equilibrium, selective sweeps do not affect density, and so the density-dependence of selection does not matter. Unfortunately this reproductive excess literature is also poorly integrated with population ecology. \cite{kimura1969natural} took constant $N$ as a requirement and then derived some variants of the logistic model that satisfy this requirement. \cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types. As a result, the role of reproductive excesses in justifying Eq.~\eqref{eq:canonical} is still largely verbal.

@@

To briefly review: based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme proposed a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) \citep{macarthur_1962}, with the latter taken to mean selection for greater saturation density \citep{gill_1974}. A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}. Setting aside $r$ which confers no selective advantage at equilibrium, we are left with $K$ and $\alpha$, which both behave like $\delta$ in that they are density-dependent and cause density to change over a sweep (although $N$ only dips temporarily during an $\alpha$-sweep). Thus, strong selection is sufficient for relative fitness to break down in the classical view of density-dependent selection. Indeed, in the defense of Eq.~\eqref{eq:canonical} given by \cite{kimura1969natural}, it was assumed that $s$ will be a few percent at most. While this may be reasonable for adaptive mutations, there is no reason to expect selection on standing variation to be so weak, wild \textit{Drosophila} being an obvious counter-example \citep{bergland_14}.

Our variable-density lottery model shows that it is not simply a lack of ecological realism that underlies the contrast between relative fitness and the classical view of density-dependent selection. Rather, in many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g. the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation). As noted in the introduction, this precludes selection concentrated at a fragile juvenile stage as a result of a reproductive excess \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}.
